
```{r setup039, include=FALSE}
source("_setupAll.R")
```

# Who went up in space for how long? {#secAstro}

"The sky's the limit." (Idiom)

**Background**  Human space travel has captured the world's attention since the first spaceflight by Yuri Gagarin\index{people!Gagarin@Yuri Gagarin} in April 1961.

**Questions**  How many men and women have gone up in space?  How long were they in space?  Which countries did they come from?

**Sources**  @tidyt:2020b

**Structure**  1277 records each of 24 variables\index{datasets!astronauts}

## What human spaceflights have there been? {#secAstro1}
The Tidytuesday project includes a collection of interesting datasets, adding a new one each week (@tidyt:2020a).  The one uploaded in July 2020 (@tidyt:2020b) contained information on manned spaceflights up to the end of 2019.  Each flight by each person is a separate record, making 1277 in all.  Several flew more than once and there are 565 different people in the dataset.  Two Americans, Jerry Ross and Franklin Chiang-Diaz, each flew seven times.  There is demographic information on the individuals (sex, year of birth, nationality) and some information on the missions (title, year, spacecraft used, length of mission).

Human spaceflight may be defined as a human reaching outer space by going beyond 100 kms from the earth.  The Challenger mission of 1986 that ended in disaster did not get that far, but is included in the dataset.  The non-fatal Soyuz mission of September 1983 when the launch vehicle caught fire before launching is not included.

\newpage

The number of people each year is shown in Figure \@ref(fig:asty).  There are sharp drops in numbers after the Soyuz 1 crash in 1967, after the Challenger disaster in 1986 and after the Columbia disaster in 2003.  The deaths in Soyuz 11 in 1971 do not seem to have had such an effect, but may have curtailed existing plans of the time.\index{histogram}

```{r asty, fig.asp = 0.5, fig.cap = 'Numbers of people by year of flight'}
data(astronauts)
ggplot(astronauts, aes(year_of_mission)) + scale_x_continuous(breaks=seq(1960, 2020, 10)) + geom_bar(fill="cyan3") + labs(x=NULL, y=NULL)
```

Early on, the participants were male and younger than later on.  There have been fewer females and they were on average younger than the males.  Figure \@ref(fig:astg) is a scatterplot\index{scatterplot} of age by year of mission with smooths\index{smooth} plotted separately for males and females.  Age has been calculated as the difference between year of mission and year of birth, so is not exact.  The vertical scale is extended beyond most of the data because of the high outlier\index{outliers}, John Glenn's\index{people!Glenn@John Glenn} third spaceflight at the age of 77 in 1998.  The males have been drawn first and the females on top to emphasise the females.\index{scatterplot}\index{smooth}

```{r astg, fig.asp=0.7, fig.cap = 'Scatterplot of age by year of flight coloured by sex, with smooths and their confidence intervals'}
astronauts <- astronauts %>% mutate(age=year_of_mission-year_of_birth)
astronauts <- astronauts %>% mutate(sex=factor(sex, levels=c("male", "female")))
astgB <- ggplot(astronauts %>% arrange(sex), aes(year_of_mission, age, group=sex)) + scale_colour_manual(values=c("springgreen3", "purple2")) + theme(legend.title=element_blank(), axis.title.y=element_text(angle=0)) + xlab("year of mission")
astgX <- astgB + geom_point(aes(colour=sex), size=1) + geom_smooth(aes(colour=sex), linewidth=2)
astgX
```

Scatterplots with many points do not convey actual numbers in groups well.  In this dataset there are 11 participants aged 41 who went up in space in 1997, 9 men and two women, including three who went up twice.  All these 14 points are represented by the same single point in Figure \@ref(fig:astg), coloured female (because of the order of plotting), although they all contribute equally to their respective smoothed fit.  Figure \@ref(fig:astb) shows barcharts\index{barchart} for males and females across the decades.  There have been far fewer females and, in particular, the bar for the only female in the 1960s, Valentina Tereshkova\index{people!Tereshkova@Valentina Tereshkova}, is barely visible.  More went into space in the 1990s than in any other decade.\index{barchart}\index{faceting}

```{r astb, out.width="80%", fig.width=6.5*0.8, fig.asp=0.7, fig.cap = 'Numbers of participants by decade and by sex'}
astronauts <- astronauts %>% mutate(decade=factor(10*(year_of_mission %/% 10)))
astbD <- ggplot(astronauts, aes(decade)) + geom_bar(aes(fill=sex)) + facet_grid(vars(sex)) + labs(x=NULL, y=NULL) + theme(legend.position="none") + scale_fill_manual(values=c("springgreen3", "purple2"))
astbD
```

## Mission times in space {#secAstro1a}
One important piece of information is the length of each mission, the time from launch until return to earth.  Figure \@ref(fig:flightt) shows a scatterplot\index{scatterplot} of this time against flight year.  Most flights were relatively short, especially in the 1960s and later on there were still many short flights of up to a few weeks, but also a group of longer ones, closer to six months, and a few very long ones of up to a year or more.  Missions with more than one person are only plotted once if all those on the mission are recorded as having the same time, so this is not a plot of individual mission times.  Missions with differing individual times are discussed later.\index{scatterplot}

```{r flightt, fig.asp=0.8, fig.cap = 'Scatterplot of mission time in hours by year of flight'}
fttX <- ggplot(astronauts, aes(year_of_mission, hours_mission)) + geom_point() + labs(x=NULL, y=NULL)
fttX
```

There is some information that might explain why flights are either short or longer, in particular whether the mission involved a visit to a space station.  That is the reason as can be seen in Figure \@ref(fig:flighttc).  A new variable has been created that records if a space station was visited and which programme it was part of: the initial U.S.S.R. Salyut programme, the US Skylab programme, the intermediate U.S.S.R. Mir programme or the ISS (International Space Station)\index{International Space Station} programme.  All the long flights are part of space station missions.  The flight in 1980 marked as "none" was in fact a space station flight by the Cuban Arnaldo Mèndez.  This is a curious error as his mission time was also incorrectly reported, multiplied by 10 in the dataset.  The case only stands out because both errors were made, otherwise it would not have been noticed.\index{scatterplot}

```{r flighttc, fig.asp=0.9, fig.cap = 'Mission time in hours by year of flight coloured by space station programme'}
astronauts <- astronauts %>% mutate(spst=factor(case_when(
              in_orbit=="ISS" ~ "ISS",
              in_orbit=="Mir" ~ "Mir",
              in_orbit=="Mir EP" ~ "Mir",
              grepl(("Skylab"), in_orbit) ~ "Skylab",
              grepl(("Salyut"), in_orbit) ~ "Salyut",
              grepl(("Saluyt"), in_orbit) ~ "Salyut",
              .default = "none"), levels=c("none", "Salyut", "Skylab", "Mir", "ISS")))
colblS <- c(colorblind_pal()(8)[c(1, 7, 3, 6, 8)])
astroF <- ggplot(astronauts, aes(year_of_mission, hours_mission)) + geom_point(aes(colour=spst)) + labs(x=NULL, y=NULL) + theme(legend.position="bottom") + theme(legend.title=element_blank())  + scale_colour_manual(values=colblS)
astroF
```

The code for this plot converted two values of the variable `in_orbit` to "Salyut", those with "Salyut" in them and those with the misspelling "Saluyt".  Minor typos like this can be irritating or seriously misleading.  The problem was discovered when colouring the display.

\newpage

### Logging mission times {#secAstro2}
As there are so many short mission times that cannot be distinguished, logarithms\index{transformation!log} of the times could be used.  Figure \@ref(fig:flighttl) shows the result.\index{scatterplot}  The vertical scale has been manually constructed to offer interpretable tickmarks.

```{r flighttl, fig.asp=0.8, fig.cap = 'Scatterplot of log mission time by year of flight coloured by space station programme'}
ftlX <- ggplot(astronauts, aes(year_of_mission, hours_mission)) + geom_point(aes(colour=spst)) + labs(x=NULL, y=NULL) + scale_y_log10(breaks=c(1, 24, 168, 365*12, 365*24),labels=c("1 hour", "1 day", "1 week", "6 months", "1 year")) + theme(legend.position="bottom") + theme(legend.title=element_blank())  + scale_colour_manual(values=colblS)
ftlX
```

There are several striking features.  The two groups of short and long missions look less variable than in Figure \@ref(fig:flightt) and the number of shorter missions declines after 2010.  The latter is a useful piece of information, as an earlier analysis concluded that mission times had increased in recent years without explaining why.  The average time certainly increased, but the reason was the decline in flights that did not visit a space station.

There are a few unexpected half points at the foot of the plot.  These are mission times recorded as zero and hence logarithms of $-\infty$.  Rather than just offering a warning message that they cannot be plotted (which **ggplot2** does in other circumstances), the semi-circles are drawn.  The 1986 value is from the Challenger disaster and the 2018 and 2019 values are for the two on an aborted 2018 mission.  For one of them, Ovchinin, the year of mission is incorrectly recorded as 2019.

There are some low outliers\index{outliers}.  The lowest one in 1986, just above one of the semi-circles, can be found by ordering\index{ordering} the dataset in ascending order of times.  It turns out that four of the seven people who died in the Challenger disaster were given times of 0 and the other three times of about 35 minutes.  It seems particularly unfortunate when the data for such tragedies are inconsistent.

Sadly, Challenger was not the only disaster.  The Space Shuttle Columbia disintegrated on re-entering the atmosphere in early 2003.  The dataset ascribes (almost) the same time to the seven on board, but the variable `in_orbit` is recorded as "Exploded" for only three of them.

The other lowish value in 1986 was for Bill Nelson, who was on Columbia's STS-61-C mission.  He was given a time of 6.03 hours, whereas five other crew members were given times of 146 hours, and one of 74 hours.  Bill Nelson's Wikipedia page (@wikiBN:2020) said he was in space for 6d 02h 03m, so perhaps the error was a result of inaccurate webscraping.  There are also three lowish values in 2005 and 2008.  These were the first Chinese flights.  Wikipedia records the first as having taken place in 2003.

Nationality is another factor to consider.  Forty nationalities are included in all, but U.S. and U.S.S.R./Russia nationalities account for `r round(100*sum(astronauts$nationality %in% c("U.S.", "U.S.S.R/Russia"))/dim(astronauts)[1])`\% of participants, so a new variable has been created for these two with the rest categorised as 'Other'.  Plotting all the cases coloured by the new variable leads to some overplotting, so faceting has been used in Figure \@ref(fig:flightn).  The whole dataset has been plotted first in grey and the groups coloured and plotted on top (ghostplotting)\index{ghostplotting}.  The Russians were first with longer flights, while the Americans took part in many more shorter flights in the 1980s and 1990s.  Other nationalities flew for the first time at the tail end of the 1970s and had more short flights.\index{scatterplot}\index{faceting}

```{r flightn, fig.asp=0.4, fig.cap = 'Scatterplots of log mission time by year of flight by nationality'}
astronauts <- astronauts %>% mutate(nation=case_when(
nationality=="U.S." ~ "U.S.",
nationality=="U.S.S.R/Russia" ~ "U.S.S.R./Russia",
.default = "Other"
), nation=fct_infreq(nation))
flx <- ggplot(astronauts %>% select(-nation), aes(year_of_mission, hours_mission)) + geom_point(colour="grey70", alpha=0.5) + labs(x=NULL, y=NULL) + scale_y_log10(breaks=c(1, 24, 168, 365*12, 365*24),labels=c("1 hour", "1 day", "1 week", "6 months", "1 year")) + geom_point(data=astronauts, aes(year_of_mission, hours_mission, colour=nation), size=2) + theme(legend.position="none") + scale_colour_manual(values=c("deepskyblue3", "red3", "darkslategray3")) + facet_wrap(vars(nation)) + theme(panel.spacing.x=unit(1, "lines"))
flx
```

## Checking the data with graphics {#secAstro4}

Checking whether other missions had differing times for their crews in the dataset uncovered a number of inconsistencies\index{inconsistencies}.  It was not enough to use the mission title, as sometimes these included multiple flights, so the dataset was grouped by the variables `mission_title`, `ascend_shuttle`, `in_orbit`, `descend_shuttle`, `year_of_mission`.  This grouped together many correctly, but failed to group all missions properly as occasionally different codings (upper/lower case, hyphens or not) were used and there were errors.  A boxplot of the biggest differences in mission times for the 538 groups found is displayed in Figure \@ref(fig:difs)\index{boxplot}.  There are four gross differences and plenty of smaller ones.  The biggest difference concerns the American/Russian trio who went to the ISS in 2001.  The time for Susan Helms is, correctly, over 160 days, while the time for James Voss is 163 hours.  Yuri Usachyov gets a similar time to Susan Helms, but is not in the same group, as the dataset reports he returned on a different spacecraft.\index{boxplot}

```{r difs, fig.asp=0.25, fig.cap = 'Boxplot of biggest differences in times for (possibly) the same missions'}
asM <- astronauts %>% group_by(mission_title, ascend_shuttle, in_orbit, descend_shuttle, year_of_mission) %>% summarise(mn=min(hours_mission), mx=max(hours_mission), dif=mx-mn)
asMg <- ggplot(asM, aes("", dif)) + geom_boxplot() + labs(x=NULL, y=NULL) + coord_flip()
asMg
```

Ideally, all typos and errors should be corrected.  The very first line of the dataset raises concern.  Yuri Gagarin\index{people!Gagarin@Yuri Gagarin}, who went up in space in April 1961, cannot have gone up in Vostock 1, been in orbit in Vostock 2, and come down in Vostock 3!  It is said of John Hartigan\index{people!Hartigan@John Hartigan}, that when he discussed datasets at Yale seminars, he might take 20 minutes to examine the first line.  This dataset is an example of why that approach can be valuable.

It would be informative to use exact flight dates, not just the years, and the dates are available on the Wikipedia page for human spaceflights (@wikiSp:2020).  The dataset illustrates some of the issues that can arise with datasets that are made publicly available and how graphics can help to identify and deal with them.  It also shows that even when there may be individual data problems, a dataset can still reveal much interesting information.

\newpage

**Answers**  Many more men than women have gone up in space and the men have been older.  Initial flights were short.  Since the space station programmes began, people have stayed much longer in space.  Most of them have been Russians or Americans, but this is changing.

**Further questions**  How many spaceflights have individuals made?  Are there differences between the groups of U.S.S.R./Russians and Americans?

### Graphical takeaways {-}
* Graphics uncover problems in datasets, be they typos, errors or more serious issues.  (§\@ref(secAstro1a) and §\@ref(secAstro4))
* Scatterplots are excellent, but it is well to check\index{checking} for duplicate pairs of values and draw additional complementary plots.\index{scatterplot}\index{duplicates}  (Figure \@ref(fig:astg))
* Colours should be used consistently for the same categories.  Different palettes should be used for other groupings to avoid confusion.\index{colour}  (Figures \@ref(fig:astg) and \@ref(fig:astb) v. Figure \@ref(fig:flighttc) or Figure \@ref(fig:flightn))
* Log (or other) transformations are effective for skewed distributions, especially with interpretable scaling.\index{transformation!log}\index{scaling}  (Figure \@ref(fig:flighttl))
* Faceting disentangles overlapping groups in scatterplots.\index{overlapping}\index{faceting}  (Figure \@ref(fig:flightn))
