
```{r setup043, include=FALSE}
source("_setupAll.R")
library(viridis)
```

# Psoriasis and the Quality of Life {#secPsor}

> "The most important practical lesson that can be given to nurses is to teach them what to observe."

> `r quote_footer('--- Florence Nightingale')`

**Background**  Psoriasis is an unpleasant skin condition and there is no known cure.  The research data discussed here come from a phase 3 clinical trial of a possible treatment.

**Questions**  Is the treatment better than the placebo?  Does it show better results on all criteria?

**Sources**  @PSI:2021a

**Structure**  Each of the 450 patients involved has two records, one for their quality of life assessment at the beginning of the study and one for the same assessment at the end of the study.\index{datasets!psoriasis}

## Comparing treatment and placebo groups {#secPsor1}

As part of their Wonderful Wednesdays project (@PSI:2021b), the Statisticians in the Pharmaceutical Industry group made available a dataset of a treatment for psoriasis (@PSI:2021a).  There was a treatment group of 300 patients and a placebo group of 150.  The PASI score (Psoriasis Area and Severity Index) was assessed for each person at the start of the study.  The DLQI questionnaire (Dermatology Life Quality Index) was administered to every patient at the start and to almost all patients after 16 weeks---a few must have dropped out.  The questionnaire has ten questions, each to be answered on a discrete scale of 0, 1, 2, 3, with 0 meaning no impact on quality of life and 3 meaning maximum impact.  Figure \@ref(fig:qlEns) is a display of three plots summarising the study.\index{spineplot}\index{scatterplot}\index{line plot}\index{ensemble of graphics}

\newpage

```{r qlEns, fig.height=5.9, fig.cap = 'The spineplot top left shows the treatment group was much bigger than the placebo group and the dropout rate was small for both groups.  The scatterplots of PASI and DLQI scores at the start of the study top right show the two groups had similar initial distributions.  The bottom plot shows that average initial scores on the 10 questions were similar for the groups (solid lines) and mostly slightly lower for the placebo group after 16 weeks (blue dashed line).  For the treatment group average scores were substantially less on all questions after 16 weeks (dark orange dashed line).'}
data(DLQI)
dq <- DLQI %>% mutate(TRTx=ifelse(TRT=="A", "Placebo", "Treatment"))
dqx <- dq %>% pivot_wider(id_cols=c(USUBJID, TRTx, PASI_BASELINE), names_from=VISIT, values_from=c(DLQI101:DLQI_SCORE))
dqx <- dqx %>% mutate(Dropout=ifelse(is.na(`DLQI_SCORE_Week 16`), "dropout", "completed"))
p1 <- ggplot(dqx) + geom_mosaic(aes(x=product(TRTx), fill=Dropout)) + theme(legend.position="none") + xlab("Group") + ylab("") + scale_fill_manual(values=c("grey70", "red"))
p2 <- ggplot(dqx, aes(PASI_BASELINE, DLQI_SCORE_Baseline)) + geom_point(aes(colour=TRTx)) + facet_wrap(vars(TRTx), strip.position="bottom") + theme(legend.position="none") + xlab("PASI at baseline") + ylab("DLQI score at baseline") + xlim(0,72) + scale_colour_manual(values=c("steelblue", "darkorange3"))
dqwS <- dqx %>% filter(Dropout=="completed") %>% group_by(TRTx) %>% summarise(across(PASI_BASELINE:`DLQI110_Week 16`, \(x) mean(x, na.rm = TRUE)))
dqws <- dqwS[, 3:22]
tdq <- data.frame(t(dqws))
colnames(tdq) <- c("A", "B")
tdq$g <- rep(c("Base", "w16"), 10)
tdq <- tdq %>% mutate(ID=rep(1:10, each=2))
p3 <- ggplot(tdq, aes(ID, A)) + geom_line(aes(linetype=g), colour="steelblue", linewidth=2) + geom_line(aes(ID, B, linetype=g), colour="darkorange3", linewidth=2) + ylim(0,3) + scale_x_continuous(breaks=seq(1,10)) + xlab("DLQI question") + ylab(NULL) + theme(legend.position="none", plot.title = element_text(vjust = 3)) + ggtitle("Average question scores by group initially (solid lines) and finally (dashed lines)")
qlEnsY <- ((p1+p2 + plot_layout(ncol=2, widths = c(1, 2)))/p3 + plot_layout(nrow=2, heights=c(1,2)))
qlEnsY
```

To be included in the study patients had to have a PASI score of at least some level.  In the dataset the minimum was 12.  Only one participant had a score over 50; the scale goes up to 72.  Some of the DLQI questions had higher averages than others, with questions 1 (how itchy, painful), 2 (how embarrassed), and 4 (influenced clothes worn) having initial averages of over 1.5.  Other questions had initial averages less than 1 or slightly over (question 5: affected social, leisure activity).

## Examining responses to individual questions {#secPsor2}

Treatment led to improvements in the averages for all the individual questions.  Averaging ordered categorical data is an approximation, so it is worthwhile looking at the separate questions in more detail.  Figure \@ref(fig:qlQ1) shows the barcharts of responses to question 1 (how itchy, painful, sore or stinging the patient's skin was over the previous week) for the two groups at the end of the trial, and compares them with how they were at the start.  The  placebo group distributions look pretty much identical, whereas the treatment group distribution moved from right to left over the trial, showing improvement for almost all patients.

```{r qlQ1, out.width="70%", fig.width=6.5*0.7, fig.cap = 'Distributions of answers to question 1 at the start of the trial (top) and at the end (bottom): quality of life declines to the right'}
ggplot(dq, aes(DLQI101)) + geom_bar(aes(fill=TRTx)) + facet_grid(rows=vars(VISIT), cols=vars(TRTx)) + ylab(NULL)  + scale_fill_manual(values=c("steelblue", "darkorange3")) + theme(legend.position="none") + xlab("DLQI Question 1 responses")
```

Of course, Figure \@ref(fig:qlQ1) does not show the changes for individual participants, just the changes in the group distributions.  Figure \@ref(fig:qlQ1t) does show the individual changes.

\newpage

The bottom row of Figure \@ref(fig:qlQ1t) shows how respondents who answered 3 to question 1 initially responded to the same question at the end of the trial.  Most of the treatment group answered 1 or 0, evidence of a lot of improvement.  Most of the placebo group answered 3 or 2, evidence of little or no improvement.  The biggest bars for each of the initial responses of 1, 2, or 3 in the placebo group are the no change bars, supporting evidence that there was little effect in the placebo group.

Figures \@ref(fig:qlQ1) and \@ref(fig:qlQ1t) have a similar form but a different meaning, so they have to be interpreted with care.

```{r qlQ1t, out.width="70%", fig.width=6.5*0.7, fig.asp=0.8, fig.cap = 'Answers to question 1 at the start of the trial (rows) and at the end (columns within the groups).'}
ggplot(dqx %>% filter(Dropout=="completed"), aes(`DLQI101_Week 16`)) + geom_bar(aes(fill=TRTx)) + facet_grid(cols=vars(fct_rev(TRTx)), rows=vars(DLQI101_Baseline), switch="y") + ylab("DLQI101 response initially") + xlab("DLQI response at Week 16") + scale_fill_manual(values=c("steelblue", "darkorange3")) + theme(legend.position="none", axis.text.y = element_blank(), axis.ticks = element_blank())
```

**Answers**  The treatment group reported better quality of life on all criteria on average.  Deeper analysis of question 1 showed that most individuals reported improvements in their quality of life.

**Further questions**  Do the answers to the other questions display similar patterns to the answers to question 1?

### Graphical takeaways {-}
* Ensembles of graphics\index{ensemble of graphics} can summarise the main features of a study succinctly.  (Figure \@ref(fig:qlEns))
* Plotting individual scores is more informative than plotting totals.  (Figure \@ref(fig:qlEns))

