
```{r setup054, include=FALSE}
source("_setupAll.R")
```

# Comparing software for facial recognition {#secFace}

> "Here's looking at you, kid."

> `r quote_footer('--- Humphrey Bogart to Ingrid Bergman in \'Casablanca\'')`

**Background**  \index{Films!Casablanca}@buol:2018 is an important paper studying how well three facial recognition\index{facial recognition} softwares performed in identifying people with different skin shades.

**Questions**  How do the error rates vary by sex and skin colour?  Are the error rates and patterns similar for the three softwares?

**Sources**  @buol:2018

**Structure**  Prediction numbers for three software packages predicting for sex across 1270 individuals with six different skin shades.\index{datasets!facial recognition}

## How well do the softwares do? {#secFace1}
The authors introduced a new facial analysis dataset, balanced by sex and skin type, that they used to test three commercial facial recognition software packages.  The study task was to determine the sex of a person from a photograph.  Results were that there was a high predictive error rate\index{Predictive error rates} for dark-skinned females (> 34\%) and an almost negligible error rate for light-skinned males (< 1\%).  The authors concluded that this unsatisfactory and uneven performance was due to the training databases the software systems used.

The summary data were reconstructed from the published paper by assuming that each percentage reported had to produce an integer number when combined with the relevant total.

\newpage

Figure \@ref(fig:faceBar) is a barchart of the numbers in each of the 12 groups (6 skin tones for each sex).  The patterns are similar, but there are more males overall, mainly in the second and sixth skin tone groups.

```{r faceBar, out.width="80%", fig.wdith=6.5*0.8, fig.asp=0.4, fig.cap = 'Numbers of different skin tones in the dataset for both sexes'}
data(aFacial)
aFacial <- aFacial %>% mutate(pred=fct_rev(Prediction), sex1=ifelse(Sex=="Female", "F", "M"), sex=fct_rev(sex1))
ggplot(aFacial, aes(Skin, weight=Freq, fill=sex)) + geom_bar() + facet_wrap(vars(sex)) + ylab(NULL) + xlab(NULL)  + scale_fill_manual(values=c("springgreen3", "purple2")) + theme(legend.position="none")
```

Figure \@ref(fig:faceh) shows the error rates by sex and skin shade (see article for details) for each of the three software systems tested using doubledecker plots\index{doubledecker plot} of error rates side by side.   Microsoft had mostly the lowest error rates---by quite a lot for the group where all software packages performed worst, the darker females.

```{r faceh, fig.asp=0.4, fig.cap = 'Error rates by sex and skin colour for three software systems drawn side by side.  In each plot males are to the left, females to the right, and skin colour gets darker from left to right within sex.  The width of each bar is proportional to the size of the group it represents.'}
hm <- ggplot(aFacial %>% filter(Software=="Microsoft")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("darkorange", "grey70")) + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Microsoft error rates")
hi <- ggplot(aFacial %>% filter(Software=="IBM")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("blue", "grey70")) + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("IBM error rates")
hf <- ggplot(aFacial %>% filter(Software=="Face++")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("lightblue", "grey70")) + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Face++ error rates")
hm + hf + hi
```

Figure \@ref(fig:face3) draws the doubledecker plots above one another to show more detail and provide a better comparison of the distributions of error rates across sex and skin tones.  The error rates were low for males and high for females, and they also increased as skin tone got darker.  Microsoft mostly did better than Face++ and did better than IBM across the board.  Males were put on the left, so that the error rates generally rose from left to right.  The softwares were ordered from overall best to worst performing from top to bottom---although none did particularly well.   The group widths are the same for each software as they are proportional to the group sizes for this criterion.

```{r face3, fig.height=7.25, fig.cap = 'Error rates by sex and skin colour for three separate software systems.  Males are on the left, females on the right, and skin colour gets darker from left to right within sex.  The width of each bar is proportional to the size of the group it represents.'}
gm <- ggplot(aFacial %>% filter(Software=="Microsoft")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("darkorange", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Microsoft error rates")
gi <- ggplot(aFacial %>% filter(Software=="IBM")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("blue", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("IBM error rates")
gf <- ggplot(aFacial %>% filter(Software=="Face++")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("lightblue", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Face++ error rates")
gm/gf/gi
```

\clearpage

## Are positive predictive values (PPV) a good criterion? {#secFace2}
There are many different possible measures for assessing binary predictions and the paper also uses PPV (Positive Predictive Value)\index{Positive Predictive Value (PPV)}.  PPV is the number of true positives divided by the total number of positives.  Ideally the predictions for each case should be compared directly, but only summary information is available. 

PPV values tend to be high.  The graphics can look more informative if (1-PPV) = FDR (False Discovery Rate)\index{False Discovery Rate (FDR)} is plotted instead.  FDRs for each combination of sex and skin shade for each software are shown in Figure \@ref(fig:faceFDR).  Whereas the bar widths in Figures \@ref(fig:face3) and \@ref(fig:faceh) were the same for each software, being proportional to the number of cases in that group, this is no longer true here.  The bar widths are proportional to the number of predictions of that group and that may be different for the different software systems.\index{doubledecker plot}

```{r faceFDR, fig.height=7.25, fig.cap = 'FDRs by sex and skin colour for three separate software systems.  Males are on the left, females on the right, and skin colour gets darker from left to right within sex.'}
aFacial <- aFacial %>% select(-Sex, -sex1, - Prediction)
aF1 <- aFacial %>% pivot_wider(names_from=pred, values_from=Freq)
aF2 <- aF1 %>% pivot_wider(names_from=sex, values_from=c(Correct, Wrong))
aF2 <- aF2 %>% mutate(freqP_M=Correct_M+Wrong_F, freqP_F=Correct_F+Wrong_M)
aF4 <- aF2 %>% pivot_longer(cols=Correct_F:freqP_F, names_to="Group", values_to="freq")
aF4 <- aF4 %>% separate(Group, into=c("Type", "Sex"))
aF4x <- aF4 %>% pivot_wider(names_from=Type, values_from=freq)
aF4y <- aF4x %>% mutate(Wrong=freqP-Correct) %>% select(-freqP)
aF4z <- aF4y %>% pivot_longer(cols=Correct:Wrong, names_to="Prediction", values_to="Freq")
aF4z <- aF4z %>% mutate(pred=fct_rev(Prediction), sex=fct_rev(Sex))
km <- ggplot(aF4z %>% filter(Software=="Microsoft")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("darkorange", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Microsoft FDR")
ki <- ggplot(aF4z %>% filter(Software=="IBM")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("blue", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("IBM FDR")
kf <- ggplot(aF4z %>% filter(Software=="Face++")) + geom_mosaic(aes(weight=Freq, x=product(pred, Skin, sex), fill=pred), divider=ddecker(), offset=0.0025) + scale_fill_manual(values=c("lightblue", "grey70")) + xlab("") + ylab("") + xlab("") + ylab("") +  scale_x_productlist(labels = paste0(rep(c("I", "II", "III", "IV", "V", "VI"), 2), "\n", c(" "," ","Male"," "," "," "," "," ","Female"," "," "," "))) + theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.background = element_blank()) + ggtitle("Face++ FDR")
km/kf/ki
```

Figure \@ref(fig:faceFDR) displays low rates, as in Figure \@ref(fig:face3), but there are important differences.  More men are predicted than there are men in the study, in particular by Face++.  The FDR values are mostly higher for males than females.  These two conclusions are related.  If you predict more males, then you will make fewer mistakes with your predictions of females.  The higher FDR rates are for skin shades V and VI, the darkest ones.

There are a number of other criteria that could be plotted and it is astonishing just how many there are, including TPR, TNR, FPR, FNR, SPC, ACC, FOR, DOR, NPV  (@PPV:2021).  The ones shown seem quite sufficient and the error rates show clearly that the three software systems are not good enough.  The analysis also shows where their weaknesses lie, with the highest error rates being for dark-skinned females.

\newpage

**Answers**  All three software packages had higher error rates for females than for males and error rates rose as skin shade darkened.

**Further questions**  If the individual data were made available, it would be informative to compare predictions across individuals and not just in summary.

### Graphical takeaways {-}
* Graphics are better than tables of percentages for displaying overall results.  (Figures \@ref(fig:face3) and \@ref(fig:faceh))
* Doubledecker plots underline why error rates can better be compared across softwares than Positive Predictive Values.  (Figures \@ref(fig:face3) and \@ref(fig:faceFDR))
