
```{r setup73, include=FALSE}
source("_setupAll.R")
```

# Varieties of plots, developing plots  {#secPlot}

> 'I repeat, as long as you have studied the theory hard enough'

> 'And what good\'s theory going to be in the real world?' said Harry loudly, his fist in the air again.

> `r quote_footer('--- Professor Dolores Umbridge and Harry Potter in \'Harry Potter and the Order of the Phoenix\'')`

## Varying various plots
There is a wide variety of plot types available (e.g., see @holtz:2018 or @schwab:2021).  Each plot type can also be drawn in many different ways: varying size, aspect ratio, scales, colour, symbols, styles, whatever.  A key message of this book is that drawing a variety of plots, and varying the plots drawn, can provide more insights than searching for a single 'optimal' display.  Just as taking many photographs of a complicated object can help people appreciate its structure, so can drawing many graphics help viewers grasp the structure of a dataset.

Approaches vary too.  Some like to study individual graphics thoroughly before moving on to other ones.  Some like to flick through lots of graphics and then return to particular ones for a closer look.  Being swamped by too many graphics can be as unsatisfactory as not having enough.  As yet there is little guidance available on how to explore large datasets effectively.

Not all graphics are the same.  There are different types of graphics---histograms, barcharts, scatterplots, and so on, there are different types of data displayed in graphics---individual cases, aggregates, statistics, parameter estimates, residuals, and so on, and there are different uses for graphics---exploration, explanation, and exposition (or presentation).  Accordingly, the same display may be interpreted in different ways depending on its context and on what kind of data it represents.

\newpage

For illustration, consider the 100 m freestyle swimming event, one which has been swum by both men and women at all Olympics since 1912.\index{datasets!Olympic Games}  Figure \@ref(fig:s100mScat) shows four scatterplots of the data.  Top left is a scatterplot of the times in seconds for the male and female gold and silver medal winners by Olympic year.  Times improved relatively steadily after initial high values.  It is likely that the females are the group of higher points and the males the group of lower ones (which could be confirmed by using colour).  Differences in gold and silver performances at later Games can barely be seen.  Top right is a scatterplot of gold medal times of women against those of men.  The winning times are strongly correlated.  Lower left is a Tukey mean-difference plot for the differences between gold and silver times for men.  There are two extreme values, one to the right for the (relatively) slow times of 1896 and one up high in 1924 when Johnny Weissmuller won by a very large margin.  Interestingly that hardly shows in the first plot.  The most surprising point is the negative value---from 1960 at Rome.  After much discussion Lance Larson was awarded the silver medal by the judges although the handtiming then used gave him a faster time than John Devitt to whom the judges awarded the gold medal.  Lower right is a residual against fitted plot for a simple linear model of women's gold medal times as a function of men's gold medal times.  The gold medal for women has been shared twice (in 1984 and 2016), so it was necessary to remove one gold medal record for each of those years.  The plot shows that a curvilinear model would give a better fit.

```{r s100mScat, fig.asp=0.6, fig.cap = 'Four different kinds of scatterplot of data for gold and silver medal performances by men and women in the Olympic freestyle 100 m'}
XGSs <- OlyX %>% filter(Event=="100m-freestyle", medalType %in% c("GOLD", "SILVER"))
XGSs <- XGSs %>% mutate(time=as.numeric(result_value)/1000)
pg1 <- ggplot(XGSs, aes(year, time)) + geom_point()
XGsmf <- unique(XGSs %>% filter(medalType=="GOLD") %>% select(year, Sex, time)) %>% pivot_wider(names_from=Sex, values_from=time)
pg2 <- ggplot(XGsmf, aes(men, women)) + geom_point() + xlim(45,65) + ylim(50,85)
XGms <- XGSs %>% filter(Sex=="men") %>% select(year, medalType, time) %>% pivot_wider(names_from=medalType, values_from=time) %>% mutate(dif=SILVER-GOLD, mm=(GOLD+SILVER)/2)
pg3 <- ggplot(XGms, aes(mm, dif)) + geom_point() + geom_hline(yintercept=0, linetype="dotted") + ylab("difference") + xlab("mean")
m1 <- lm(data=XGsmf, women~men)
XGr <- XGsmf %>% filter(year>1908)
XGr[, "residuals"] <- m1$residuals
XGr[, "fitted"] <- m1$fitted
pg4 <- ggplot(XGr, aes(fitted, residuals)) + geom_point() + geom_hline(yintercept=0, linetype="dotted")
(pg1 + pg2)/(pg3+pg4)
```

\newpage

A set of boxplots in parallel could be plots of individual variables or plots of one variable conditioned on another.  Figure \@ref(fig:finch2box) displays two groups of boxplots.  Above, there are boxplots of the five beak measurements for Darwin's finches in the Isabela Island dataset (Chapter \@ref(secFinches)).\index{datasets!Darwin's finches}  All are skewed\index{skewness} to the right, especially beak height and width, doubtless due to the mix of species.  Below, there are boxplots of beak height for the five species on Isabela Island.  Two have clearly bigger beak heights than the other three.  It is important to indicate what kind of plot each is.  Titles and colour can help (intentionally left out here), as can using boxplot widths to show group sizes.  Barring missing values (there are none here) the boxplots are based on equal numbers of cases in the first plot.  In the second the numbers vary from 11 *fortis platyrhyncha* to 81 *fuliginosa parvula*.

```{r finch2box, fi.asp=0.6, fig.cap = 'Boxplots of beak measurements (above) and beak height by species (below) on Isabela Island'}
fp <- mSH1 %>% select(BeakW:N.UBkL) %>% pivot_longer(cols=BeakW:N.UBkL, names_to="Var", values_to="Val")
fpx <- ggplot(fp, aes(Var, Val)) + geom_boxplot() + ylab(NULL) + xlab(NULL) + coord_flip()
fpy <- ggplot(mSH1, aes(species, BeakH)) + xlab(NULL) + ylab(NULL) + geom_boxplot() + coord_flip()
fpx/fpy
```

The individual graphics of Figures \@ref(fig:s100mScat) and \@ref(fig:finch2box) have purposely been drawn similarly.  In practice groups of graphics like this would be drawn distinctly.  For instance, whatever else one might think of them, the four graphics standardly produced by plotting a model in R are.  The danger lies more in glancing quickly at isolated, individual graphics and not checking\index{checking} what they display.

\newpage

## Ensembles of plots
Thanks to trellis plots (@becker:1996), lattice plots (@sarkar:2008), and faceting\index{faceting} (@wilkinson:1999 and @wickham:2016), displays of small multiples\index{small multiples} are well understood.  Each of the component plots is of the same type and is conditioned on other dataset variables.  By comparison, ensembles made up of different types of plots have been less discussed.  @unwin:2018 suggest that the key element is coherence: having a coherent theme, a coherent alignment, coherent look, consistent scales and formatting.  Examples include Figures \@ref(fig:movieRunYB) and \@ref(fig:uefaTQ), and the set of four Figures \@ref(fig:PalEns0) to \@ref(fig:PalEns3).  Another example is Figure \@ref(fig:comMFmed), repeated here as Figure \@ref(fig:comMFmedX).

```{r comMFmedX, fig.cap = 'Median times for men and women for "up" (orange) and "down" (black) races of the Comrades Marathon since 1980'}
CM12 + plot_layout(heights=c(1.2, 0.4))
```

The graphics are aligned horizontally and vertically, as are the scales.  The colours above refer to the "up" and "down" races, while the colours below refer to the sex of the runners.

\newpage

A more complex ensemble is shown in Figure \@ref(fig:qlEns) reproduced here as Figure \@ref(fig:qlEnsX).  Three separate pieces of information are compared for the two groups with three different sets of graphics: proportions of dropouts in a barchart, two initial measurements in scatterplots, ten life quality index question means with line charts.  There is no need for common scaling as the variable scales all differ.  The groups are coloured alike in the scatterplots and the line charts, but not in the barchart, where the dropouts are coloured to emphasise how few there were and how similar the proportions were.  The lines are drawn differently according to whether the responses were at the start or end of the study.


```{r qlEnsX, fig.cap = 'The spineplot top left shows that the treatment group was much bigger than the placebo group and that the dropout rate was small, about the same for both groups.  The scatterplots of PASI scores and DLQI scores at the start of the study top right show that the two groups had similar initial distributions.  The bottom plot shows that the average initial scores on each of the 10 questions were similar for the two groups (solid lines) and mostly slightly lower for the placebo group after 16 weeks (blue dashed line).  For the treatment group the average scores were substantially less on all questions (dark orange dashed line)'}
qlEnsY
```

\newpage

## Differences between early drafts and later graphics

Quickly drawn default graphics can reveal valuable information.  Taking the time to improve them can make it easier to see this information by making it stand out and can uncover further information.  It is instructive to compare pairs of initial and better graphics to study what changes might be made and how effective they are.  Readers will have different opinions on what might be done and how.  They are welcome to experiment with the data themselves.

The term "initial" here usually means that the graphic has been drawn after some data cleaning.  Quite often the first graphics drawn indicate further data problems of one kind or another that require data wrangling to fix.  So there can be many early graphics, just as there can be many later graphics that are amended in a range of ways.  What there cannot be is a final graphic; there are always potential alternatives.  And whatever improvements can be made to individual graphics, most information will be gathered by looking at more than one data display---however good an individual graphic is.

### 1912 Democratic Convention
Chapter \@ref(sec1912Dem) studied the selection of the Democratic candidate for President at the 1912 Convention.\index{datasets!Democratic Convention 1912}  The voting was initially shown in Figure \@ref(fig:candvs) and later, with changes, in Figure \@ref(fig:candvs2).  The two plots are shown here in Figures \@ref(fig:candvsX) and \@ref(fig:candvsX2).  The changes made are listed below.

```{r candvsX, fig.cap = 'Voting at the 1912 Democratic convention shown in the initial plot'}
cand1
```

```{r candvsX2, fig.asp=0.5, fig.cap = 'Voting at the 1912 Democratic convention shown in the final plot'}
cand2
```

The main features seen in the first plot were that there were many candidates, but only a few serious ones; that Clark was ahead initially, particularly after the 10th ballot; that Wilson gained steadily and made the decisive breakthrough around the 43rd ballot.  The second plot emphasises the key changes and shows that the breakthrough for Wilson took place on the ballot after the last adjournment.

Changes include:

1. The simple order of ballots on the horizontal scale has been replaced by estimated ballot times.  
2. Only candidates who received over 35 votes in at least one ballot have been included.  This was to concentrate on the main candidates and reduce the clutter due to minor candidates who never had much support.  The legend in Figure \@ref(fig:candvsX) listed 13 candidates and the category "Notvoting".  Figure \@ref(fig:candvsX2) displays the voting for only 5 candidates.  
3. The plot background has been coloured white when the convention was in session and grey when it was not.  
4. The aspect ratio of the plot has been changed to 0.5.  Time series are generally better wider than tall, not square or taller than wide.  
5. The dataset has been expanded to include ballots in which the candidates got zero votes.  Otherwise the lines for candidates who withdrew were left hanging and the connecting lines for candidates who had no support for some intermediate ballots were not correctly drawn (as happened in Figure \@ref(fig:candvsX) for Harmon).  
6. Vote changes have been shown by vertical lines when the ballot occurred, not as a sloping line between consecutive ballots. This reflects what actually happened.  A disadvantage is where lines are drawn on top of one another, as at the last ballot.  
5. Points have been drawn for each ballot to emphasise when and how often they took place, especially when very little changed between ballots.  
6. Horizontal lines have been added at two-thirds support (needed to win the nomination) and at 50\% support.  Both levels played important parts in the convention.  (At the Republican convention a simple majority sufficed.)  
7. A colour blind scale has been used instead of the default colour scale\index{default scaling}.  Colour blind scales assist in distinguishing the candidates.  
8. Candidates have been plotted in reverse order of their votes in the penultimate ballot so that if two candidates overlap the colour of the candidate with more votes is on top.  
9. The legend order has been changed to match the order in the last few ballots.  This makes it easier to identify the lines for each candidate.  
10. The legend title and the vertical axis label have been removed.  These texts are not necessary and take up space and attention.

### Gapminder Life Expectancies
The first plot in Chapter \@ref(secGapm), Figure \@ref(fig:LEyr) showed how life expectancy changed for many countries round the world over the past 200 years.  Several insights could be derived, but the display was overcrowded and there might have been more information in the data.  Figure \@ref(fig:LEf4) was an alternative version with the countries faceted and coloured by region.   Figure \@ref(fig:gapLEX) displays both figures.\index{datasets!Gapminder}  

```{r gapLEX, out.width="45%", fig.show="hold", fig.cap = 'Gapminder life expectancy data drawn for all countries (left) and with countries grouped by region (right)'}
pgmX
pr4X
```

The initial plot shows the steady increase in life expectancy in the 20th century, the surprisingly constant values for many countries in the 19th century, and occasional sharp falls and rises for individual countries.  The later plot shows that life expectancy was generally higher in Europe than elsewhere in the 19th century, that the overall patterns, if not the levels, were similar in all four regions, and that there were countries whose life expectancy graphs differed strikingly from others in their region.

Changes include:

1. The countries have been coloured by region to provide some information on geographic location.\index{colour}  
2. Each region has been plotted separately so that regional differences and local outliers can be seen.\index{faceting}  
3. Some 13 small countries with incomplete data were excluded as they provided little information.

\newpage

### Electric car charging
Chapter \@ref(secElecC) included two graphics of the use of charging stations over time, Figure \@ref(fig:elecStat) and Figure \@ref(fig:elecStat2).  They are drawn here as side by side in smaller form in Figure \@ref(fig:elecStat12). \index{dotplot}\index{datasets!electric car charging}

```{r elecStat12, fig.asp=1.2, fig.cap = 'When charging stations were used'}
plan1 <-"
  ##BBB
  AABBB
  AABBB
  "
ElecSt1 + ElecSt2 + plot_layout(design=plan1)
```

From the first plot it is apparent that a number of charging stations only started to be used late on in the study, and that some charging stations were used relatively little.  The second plot highlights when the testing phase stopped.  It shows that some charging locations were little used, that office locations came later than R&D locations, that there were differences worth investigating further in patterns of use of charging stations at the same location.

Changes include:

1. Using a high aspect ratio\index{aspect ratio} for the amended plot in Figure \@ref(fig:elecStat2) to show the points for the individual charging stations better.  
2. Sorting charging stations by location to be able to see how usage of neighbouring stations compares.\index{sorting}  
3. Ordering locations by first use to put them in an order matching the development of the study.\index{ordering}  
4. Ordering charging stations within location to show how the installations developed.  
5. Colouring facility types to highlight group differences.  
6. Drawing a dashed red line at the time the study switched from the test phase to the analysis phase.  This emphasises how many locations were not used (not available?) during the test phase.  
7. Dropping the vertical axis label and the individual Station IDs.  The caption makes clear what is shown vertically and the Station IDs are just codes that provide little information.  
8. Adding Location ID labels.  Although the labels are uninterpretable codes, they do emphasise that the locations are different.  Should someone have access to which locations those codes relate to, they could look up the details.  
9. The date formatting in the default plot is does not include the year and does not explain where in a month the month's name is placed.  Detailed dates provide better context.  The horizontal axis label was dropped as the name was uninformative and it is obvious from the more detailed labelling that time is being plotted.\index{dates}

Additionally a much fuller caption was used in the amended plot in Chapter \@ref(secElecC).

### Gay rights
The choropleth map in Chapter \@ref(secGayR), Figure \@ref(fig:gayMap) gave an overview of the geographic distribution of support for same-sex marriage at state level.  Individual states have very different population sizes and the survey design reflected that.  A default plot of confidence intervals for the state estimates is shown here with the amended plot drawn in the chapter, Figure \@ref(fig:gayCIs), in Figure \@ref(fig:gayCIX).\index{datasets!Annenberg election survey 2004}

```{r gayCIX, fig.asp=1.2, fig.cap = 'Support for same-sex marriage by state with 95\\% confidence intervals'}
gayCId <- ggplot(USpsl, aes(NAME, p1)) + geom_point() + geom_errorbar(aes(ymin = low, ymax = upp)) + coord_flip()
gayCId + gayCIx
```

The first plot shows that states have different rates and different confidence interval lengths.  Many pairs of confidence intervals do not overlap.  The second plot shows that: the few states with over 50% support are in the North East; Washington DC and Delaware have rates more like the North East than the South; MidWest states have similar rates to each other; the Dakotas have broader intervals (i.e. smaller survey samples matching smaller populations).

Changes include:

1. The states have been sorted by Census Bureau region so that local comparisons are easier.\index{sorting}  
2. States have been ordered by estimate within region, again to facilitate comparisons.\index{ordering}  
3. Estimates have been coloured by region to differentiate the groups.\index{colour}  
4. Points have been drawn bigger to make the estimates themselves stand out more.  
5. Intervals have been drawn as straight lines without end bars to reduce clutter and give less emphasis to the smaller states.  
6. Horizontal scaling has been changed to percentages instead of proportions to better match the data.  
7. Horizontal limits have been set to [0,100%] to avoid false impressions of the amount of support.\index{scaling}  
8. Breakpoints on the horizontal axis have been chosen to include the critical 50% value, so that it is easy to see which intervals do not even reach majority support.  
9. Axis name labels have been dropped to leave more space for the graphic.

### Human space flight
Figure \@ref(fig:flightt) in Chapter \@ref(secAstro) shows the mission time of space flights by year of flight.  If the flights went to a space station, they were coloured accordingly in Figure \@ref(fig:flighttc).  The final version after taking logarithms was given in Figure \@ref(fig:flighttl).  Figures \@ref(fig:flightt) and Figure \@ref(fig:flighttl) are redrawn here as Figure \@ref(fig:flighttz).\index{datasets!astronauts}

```{r flighttz, fig.asp=0.4, fig.cap = 'Mission times by year (left) and log mission times by year coloured by space station programme (right)'}
fttX + ftlX
```

The plot on the left shows that most mission times were short, especially before 1980, that there were a few very long missions, and that there was a group of missions between about 4000 and 5000 minutes.  The plot on the right show that mostly the missions split into short ones of a week or so, that there were longer ones of several months as part of space station programmes, that there were some low mission times that required further investigation.

Changes include:

1. The logarithms of mission times have been used instead of the mission times as the plot then shows more structure.\index{transformation!log}  (Figure \@ref(fig:flighttc) revealed that it primarily splits the missions into whether they were to a space station or not.)  
2. The breaks on the vertical axis scale were not set equally but chosen to reflect familiar time periods to make the times readily understandable.\index{scaling}  
3. The display was coloured by space station programme to explore the differences between the shorter and longer missions.\index{colour}

\newpage

### Euro 2020
For each team taking part in the Euro 2020 soccer competition, Figure \@ref(fig:uefa20) showed in which countries the players in the squad played.  A default plot for this example is shown in Figure \@ref(fig:uefa20X).  The figure from Chapter \@ref(secEuros) is replotted below.\index{datasets!Euros teams}

```{r uefa20X, fig.asp=1.2, fig.cap = 'Default plot of numbers of Euro 2020 players of each team playing in other countries\' leagues'}
uefa20y <- ggplot(eu20p %>% filter(year=="2020"), aes(club_country)) + geom_bar(aes(fill=club_country)) + facet_wrap(vars(nat_team)) + coord_flip()
uefa20y
```

```{r uefa20Y, fig.asp=1.2, fig.cap = 'Amended plot of numbers of Euro 2020 players of each team playing in other countries\' leagues'}
uefa20w
```

The first plot shows that some teams' players are from only a few countries' leagues and some from several.  In the second plot the teams and leagues can be seen and individual patterns identified (e.g., England, Italy, Russia, Spain, Austria, Wales).

Changes include:

1. The teams have been ordered by the percentage of players playing in their own national league.\index{ordering}  
2. The bars have been ordered\index{ordering} by the total number of players from that league playing in Euro 2020.  Countries with less than 9 players have been grouped into "Other".  This emphasises the more important leagues and makes the labels readable.  
3. Colours close to country colours have been used instead of default colours.\index{colour}  
4. The legend has been dropped as it takes up a lot of room (even with fewer countries) and the labels show what the colours mean.\index{legend}  
5. The number of rows has been set to 4 to get a better balanced display and give more vertical space for the country labels of the leagues.  
6. Axis title labels have been dropped leaving more space for the display.  
7. The horizontal axis scale has been dropped as exact counts are not important.  Extra space has been added to the ends of the scale to improve readability.\index{scaling}  
8. Gridlines have been removed to reduce clutter.

\newpage

### Characteristic changes
1. Grouping information supports understanding.  It is easier to work with chunks of information than with all the individual components.  Grouping may be achieved by spatial separation (Figure \@ref(fig:elecStat12)), by faceting (Figures \@ref(fig:gapLEX) and \@ref(fig:gayCIX)), by colour (Figures \@ref(fig:elecStat12) and \@ref(fig:flighttz)).  
2. Plot aspect ratios are important (Figures \@ref(fig:candvsX2) and \@ref(fig:elecStat12)).  
3. Ordering is an effective way of emphasising patterns (Figures \@ref(fig:gayCIX) and \@ref(fig:uefa20Y)).  
4. Colour is a powerful tool and should be chosen with care (Figures \@ref(fig:candvsX2), \@ref(fig:gapLEX), and \@ref(fig:flighttz)).  
5. Reducing clutter and making space by reformatting or by excluding less important cases is always worth considering (Figure \@ref(fig:gapLEX) and Figure \@ref(fig:uefa20Y)).  
6. Scaling and labelling should be sufficient but not excessive (Figures \@ref(fig:gayCIX) and \@ref(fig:uefa20Y)).  
7. Time scaling and labelling should be chosen carefully (Figures \@ref(fig:candvsX2) and \@ref(fig:flighttz)).  
8. Adding guidelines for critical values is useful (Figures \@ref(fig:candvsX2) and \@ref(fig:elecStat12)).    

## Classifying graphics
Graphics may be classified in different ways and a number of terms are used.  Data graphics generally refers to displays of raw data.  Model graphics implies displays associated with models, such as plots of predictions or residuals, plots of parameter estimates or of statistics for comparing models.  Multivariate graphics includes displays showing many variables such as parallel coordinate plots\index{parallel coordinate plot} and mosaicplots\index{mosaicplot}, small multiples\index{small multiples} including facetings\index{faceting}, and dimension reduction displays such as biplots and multidimensional scaling plots.  Interactive graphics\index{interactive graphics} describes displays which can be directly manipulated: queried, sorted, reformatted, linked across multiple displays.  Dynamic graphics\index{dynamic graphics} means displays which can be set in motion, such as rotating plots and animated plots.  There is some overlap between interactive and dynamic graphics, but the emphasis with the former is on interaction and the emphasis with the latter is on dynamics.  And there are further terms mentioned in the literature.

It is not just the technical aspects and the data background that affect the use of graphics.  The setting is important.  This can range from a publication of a research project in a specialist field to an item on the television news about the progress of an epidemic.  In some circumstances a graphic may be paid close attention, in others the graphic may be competing for attention.  Sometimes much must be explained, sometimes it is safe to assume substantive knowledge on the part of the graphic's viewers.  Graphics are not an easy matter.

### Main points {-}
* Not all graphics that look the same are the same.  Context matters.
* Ensembles of plots require coherence of look, theme, alignment, scales, and formatting.
* Graphics can be changed in many ways and describing why a change has been made is more enlightening than describing how it has been made.



